---
title       : Are your hands in proportion with your height?
subtitle    : A quick tour
author      : Hadrian84
job         : R-enthousiast
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : solarized_light 
widgets     : [bootstrap, quiz]        # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
logo        : Rlogo.png
github:
   user: Hadrian84
   repo: DatascienceAssignment
---

## Introduction

This application asks for four inputs:

- Your name
- Gender
- Height
- Hand length 

It also relies on a dataset containing many body statistics. This dataset is publicly available (ANSUR: [click here](http://mreed.umtri.umich.edu/mreed/downloads.html#ansur)).

For the purpose of this tour, we used randomized data.

---

## First graphs generated

```{r, echo = FALSE, fig.height = 5, fig.align = 'left'}
library(ggplot2)
library(gridExtra)

set.seed(123)
xs  = sort(rnorm(10000, 174, 3))
ys  = sort(rnorm(10000, 190, 4))
yss = ys + rnorm(10000, 0, 4)
dat = data.frame(value = 1:10000, x = xs, y = ys, regression = yss)

height = 174
hand   = 190

g1 = ggplot() +
     geom_density(data = dat, aes(xs), fill = "steelblue4", color = "steelblue4", alpha = 1/2) +
     scale_y_continuous(limits = c(0, 0.15)) +
     theme(legend.position = "null") +
     labs(x = "Body Height (cm)", title = "Body height distribution") +
     geom_vline(xintercept = height, color = "black", size = 1.1) +
     annotate("text", x = height - 1, y = 0, angle = 90, hjust = 0, color = "black", label = "Hadrian84") +
     annotate("text", x = height, y = 0.075, label = "50%", 
              color = "white", size = 8)     

g2 = ggplot() +
     geom_density(data = dat, aes(ys), fill = "steelblue4", color = "steelblue4", alpha = 1/2) +
     scale_y_continuous(limits = c(0, 0.105)) +
     theme(legend.position = "null") +
     labs(x = "Hand Length (mm)", title = "Hand length distribution") +
     geom_vline(xintercept = hand, color = "black", size = 1.1) +
     annotate("text", x = hand - 1, y = 0, angle = 90, hjust = 0, color = "black", label = "Hadrian84") +
     annotate("text", x = hand, y = 0.05, label = "50%", 
              color = "white", size = 8)     

grid.arrange(g1, g2, ncol = 2)

```

Based on your input, it compares your **height** and **hand length** with a representative population. 

It also shows what percentage of people have higher values than you.

--- 

## Summarizing graph

```{r, echo = FALSE, fig.height = 5, fig.align = "left"}
g3 = ggplot(dat, aes(x = xs, y = regression)) +
     geom_point() +
     geom_smooth(method = "lm", se = FALSE, size = 1.5, linetype = "dashed", color = "blue") +
     geom_point(aes(x = height, y = hand), color = "red",  cex = 2.5) +
     geom_vline(xintercept = height, color = "red", linetype = "dashed", alpha = 1/2) +
     geom_hline(yintercept = hand, color = "red", linetype = "dashed", alpha = 1/2) +
     labs(x = "Body Height (cm)", y = "Hand Length (mm)", title = "Relationship body height and hand length")

print(g3)
```

This regression plot shows the overall **height** and **hand length** relationship (dotted blue line). As expected, taller people have bigger hands (and vice versa). 

Where you reside in this plot is indicated by the red dot.

--- 

## Concluding graph

```{r, echo = FALSE, fig.height = 5, fig.align = "left"}
g4 = ggplot() +
     geom_density(data = dat, aes(ys), fill = "steelblue4", color = "steelblue4", alpha = 1/2) +
     scale_y_continuous(limits = c(0, 0.105)) +
     theme(legend.position = "null") +
     labs(x = "Hand Length (mm)", title = "Hand length (given Height)") +
     geom_vline(xintercept = hand, color = "black", size = 1.1) +
     annotate("text", x = hand - 1, y = 0, angle = 90, hjust = 0, color = "black", label = "Hadrian84") +
     annotate("text", x = hand, y = 0.05, label = "50%", 
              color = "white", size = 8)  

print(g4)
```

The final graph uses a segment of the regression plot and displays your **hand length** in the hand length distribution of people with your height. 

Again, it shows what percentage of people (with your height) have bigger hands than you.